---
title: "Differential Expression"
output: html_document
---


Use this notebook to perform differential expression and all other post-DE
analyses.

importing needed libraries.
``` {r}
library('tidyverse')
library('DESeq2')
library('fgsea')
library('ggplot2')
library('gridExtra')
```

functions to make counts matrix and colData for DESeq
```{r}
make_counts <- function(filename){
  # read in counts matrix
  counts <- read_csv(filename)

  rowData <- counts['gene']
  
  counts_matrix <- as.matrix(counts[, -1])
  rownames(counts_matrix) <- rowData$gene
  return(counts_matrix)
}

make_colData <- function(counts) {
  #make coldata
  coldata <- data.frame(
    samples = colnames(counts), 
    time = c(rep('P0', 2), rep('P4', 2), rep('P7', 2),rep('AD',2)),
    row.names='samples')
  coldata$time <- as.factor(coldata$time)
  coldata$time <- relevel(coldata$time, ref='P0')
  
  return(coldata)
}
```

reading in counts matrix data and making colData
```{r}
# make counts matrix
counts_matrix <- make_counts('results/full_verse_concat_filtered.csv')

#make column data matrix
coldata <- make_colData(counts_matrix)
```

Running DESeq using DESeqDataSetFromMatrix to compare timepoints
```{r}
# Run DESeq, comparing p0 and AD timepoints and generate a tibble containing the results
dds <- DESeqDataSetFromMatrix(countData = counts_matrix, colData = coldata, design = ~time)
dds <- DESeq(dds)
res <- results(dds, contrast=c('time', 'AD', 'P0'))

  # order results by pval
resOrdered <- res[order(res$pvalue),] %>% 
  as_tibble(rownames='geneids')

# read in id2gene file
id2gene <- read.table('results/full_id2gene.txt')
names(id2gene) <- c("geneids", "genenames")

# Join the gene names into the results tibble while preserving all of the original information in the results
resWGeneid <- resOrdered %>% 
  left_join(id2gene, by='geneids') %>% 
  select(geneids, genenames, padj, log2FoldChange)
```


FGSEA ANALYSIS
```{r}
make_ranked_log2fc <- function(gene_results) {
  gene_results_no_NA <- drop_na(gene_results)
  
  #order genes based on log2FoldChange
  ranked_genes <- as_tibble(gene_results_no_NA) %>%
    arrange(desc(log2FoldChange)) %>% 
    dplyr::select(genenames, log2FoldChange)
  
  # Create a list with symbol and log2FoldChange values
  ranked_genes_vec <- deframe(ranked_genes)
  
  return(ranked_genes_vec)
}

# make ranked gene list
rnk_list <- make_ranked_log2fc(resWGeneid)

# run fgsea
gene_set <- fgsea::gmtPathways('m2.all.v2023.2.Mm.symbols.gmt')
fgsea_results <- fgsea(gene_set, rnk_list, minSize = 15, maxSize= 500)
fgsea_results <- fgsea_results %>% as_tibble()
```

recreate figure 1D - D) expression values of representative sarcomere, mitochondrial, and cell cycle genes significantly differentially expressed during in vivo maturation (P0, P4, P7, AD).
``` {r}
#investigating 
top_pathways <- function(fgsea_results, num_paths){
  fgsea_results <- fgsea_results %>% mutate(
    NES_plot_status = ifelse(
      NES < 0, "neg", "pos")
  )
  top_ten_pathways <- fgsea_results %>% 
    arrange(NES)%>%
    head(num_paths)
  bottom_ten_pathways <- fgsea_results %>% 
    arrange(NES)%>%
    tail(num_paths)
  pathways <- rbind(top_ten_pathways, bottom_ten_pathways)
}
pathways <- top_pathways (fgsea_results, 50)
View(pathways)
```

Extracting norm counts
```{r}
normalized_counts <- counts(dds, normalized = TRUE)
print(normalized_counts)

```
their plot has P0, P4, P7 and AD timepoints -> average of each time point for each gene

```{r}

time_points <- sub("rep\\d+", "", colnames(normalized_counts))

# Create a new matrix to store averaged counts
averaged_norm_counts <- matrix(NA, nrow = nrow(normalized_counts), ncol = length(unique(time_points)))
colnames(averaged_norm_counts) <- unique(time_points)
rownames(averaged_norm_counts) <- rownames(normalized_counts)

# Calculate average counts for each time point
for (time_point in unique(time_points)) {
  time_point_columns <- grep(time_point, colnames(normalized_counts))
  averaged_norm_counts[, time_point] <- rowMeans(normalized_counts[, time_point_columns])
}

# Print the averaged counts matrix
print(averaged_norm_counts)
```
Adding gene name by gene id
```{r}
avg_norm_counts_genes <- averaged_norm_counts %>% 
  as_tibble(rownames='geneids') %>%
  left_join(id2gene, by='geneids')

avg_norm_counts_genes <- avg_norm_counts_genes[, -1] %>%
  select(genenames, everything())

print(avg_norm_counts_genes)

```

Isolating counts 
list from paper 
  Mitochondria: Mpc1, Prdx3, Acat1, Echs1, Slc25a11, Phyh
  Cell cycle: Cdc7, E2f8, Cdk7, Cdc26, Cdc6, E2f1, Cdc27, Bora, Cdc45, Rad51, Aurkb, Cdc23
  Sarcomere: Pdlim5, Pygm, Myoz2, Des, Csrp3, Tcap, Cryab
  
```{r}
#filter norm counts results for mitochondrial genes
mitochon_genes = c('Mpc1', 'Prdx3', 'Acat1', 'Echs1', 'Slc25a11', 'Phyh')
mitochon_counts_results <- avg_norm_counts_genes %>%
  filter(genenames %in% mitochon_genes)

print(mitochon_counts_results)

#filter norma counts results for cell cycle genes
cell_cycle_genes = c('Cdc7', 'E2f8', 'Cdk7', 'Cdc26', 'Cdc6', 'E2f1', 'Cdc27', 'Bora', 'Cdc45', 'Rad51', 'Aurkb', 'Cdc23')
cell_cycle_counts_results <- avg_norm_counts_genes %>%
  filter(genenames %in% cell_cycle_genes)

print(cell_cycle_counts_results)

#filter norm counts results for sarcomere genes
sarcomere_genes = c('Pdlim5', 'Pygm', 'Myoz2', 'Des', 'Csrp3', 'Tcap', 'Cryab')
sarcomere_counts_results <- avg_norm_counts_genes %>%
  filter(genenames %in% sarcomere_genes)

print(sarcomere_counts_results)

```
plotting
```{r}
#need to pivot longer
mitochon_counts_results_longer <- mitochon_counts_results %>%
  pivot_longer(cols = -genenames, names_to = "time_point", values_to = "normalized_count")
print(mitochon_counts_results_longer)

cell_cycle_counts_results_longer <- cell_cycle_counts_results %>%
  pivot_longer(cols = -genenames, names_to = "time_point", values_to = "normalized_count")
print(cell_cycle_counts_results_longer)

sarcomere_counts_results_longer <- sarcomere_counts_results %>%
  pivot_longer(cols = -genenames, names_to = "time_point", values_to = "normalized_count")
print(sarcomere_counts_results_longer)
```
```{r}
# Plot using ggplot2
# Define numeric values for time points
time_point_numeric <- c(P0 = 1, P4 = 2, P7 = 3, AD = 4)

# Convert time_point to numeric based on the defined values
mitochon_counts_results_longer <- mitochon_counts_results_longer %>%
  mutate(time_point_numeric = time_point %>% factor(levels = c("P0", "P4", "P7", "AD")) %>% as.numeric())

# Convert time_point to numeric based on the defined values
cell_cycle_counts_results_longer <- cell_cycle_counts_results_longer %>%
  mutate(time_point_numeric = time_point %>% factor(levels = c("P0", "P4", "P7", "AD")) %>% as.numeric())

# Convert time_point to numeric based on the defined values
sarcomere_counts_results_longer <- sarcomere_counts_results_longer %>%
  mutate(time_point_numeric = time_point %>% factor(levels = c("P0", "P4", "P7", "AD")) %>% as.numeric())

# Set fixed margins for plot titles
theme_title_fixed <- function(title_size = 12, margin_top = 0.5, margin_bottom = 0.5) {
  theme(
    plot.title = element_text(size = title_size, margin = margin(margin_top, 0, margin_bottom, 0))
  )
}

# Plot with ordered time points

plot1 <- ggplot(data = sarcomere_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Sarcomere") +
  theme_minimal() + 
  theme(legend.key.size = unit(0.3, "cm"),
        legend.spacing.x = unit(0.05, "cm"),  # Adjust horizontal spacing
        legend.spacing.y = unit(0.05, "cm"))   # Adjust vertical spacing)


plot2 <- ggplot(data = mitochon_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Mitochondria") +
  theme_minimal()+ 
  theme(legend.key.size = unit(0.3, "cm"),
        legend.spacing.x = unit(0.05, "cm"),  # Adjust horizontal spacing
        legend.spacing.y = unit(0.05, "cm"))

plot3 <- ggplot(data = cell_cycle_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Cell Cycle") +
  theme_minimal()+ 
  theme(legend.key.size = unit(0.05, "cm"),
        legend.spacing.x = unit(0.05, "cm"),  # Adjust horizontal spacing
        legend.spacing.y = unit(0.05, "cm"))

# Arrange plots vertically
grid.arrange(plot1, plot2, plot3, ncol = 1)

```
individual plots


```{r}

ggplot(data = sarcomere_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Sarcomere") +
  theme_minimal()


ggplot(data = mitochon_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Mitochondria") +
  theme_minimal()

ggplot(data = cell_cycle_counts_results_longer, aes(x = time_point_numeric, y = normalized_count, color = genenames)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = c(1, 2, 3, 4), labels = c("P0", "P4", "P7", "AD")) +  # Set x-axis labels
  labs(x = "Time Point", y = "Normalized Count", color = "Gene", title= "Cell Cycle") +
  theme_minimal()

```
